\Dchapter{Related work\either{ to Automatic Annotations}{}}

%The field of dynamic analysis has a rich history.
%Ball~\cite{ball1999concept}
%introduces frequency spectrum analysis,
%an approach that observes a running program
%that is similar to our instrumentation approach.
%Mock~\cite{mock2003dynamic}
%makes the case for efficient profiling of programs
%to better facilitate usage of instrumentation.
%
%Value Profiling is another related area which characterises
%programs based on their running entities.
%
%Daikon~\cite{ernst2001dynamically}
%uses dynamic analysis to recover likely program invariants
%in C programs.
%
%Dynamic type inference has been attempted in many different
%areas.
%Rubydust~\cite{An10dynamicinference}
%infers static types for Ruby. They prove the generated types
%sound, which we do not. 
%\begin{verbatim}
%Conversely, they do not generate
%recursive types, but recursive types in ruby are probably
%nominal, so how different are we?
%\end{verbatim}
%
%In the context of JavaScript, several usages of this technique
%can be found.
%JSTrace~\cite{saftoiu2010jstrace}
%generates types for (?).
%Separately, work has been done to generate JSDoc-like annotations~\cite{odgaard2014}.
%TypeDevil~\cite{pradel2015typedevil}
%uses dynamic analysis to warn JavaScript programmers of possible inconsistencies
%in their programs.
%
%Work in recovering context-free grammars is most related to our algorithm
%to recover recursive types.
%% TODO Shamir~\cite{shamir1962remark} notes that it is impossible
%% TODO \cite{knobe1976method}
%
%In the context of machine learning, 
%this area is called grammar induction or language learning.  % according to vcrepinvsek2005inferring
%% TODO Wang\cite{wang1998grammar} summarises 
%{\v{C}}repin{\v{s}}ek et. al~\cite{vcrepinvsek2005inferring}
%use genetic programming to infer context-free grammars
%for domain-specific languages.
%Most work in this area assume both positive and negative
%examples. We cannot distinguish between these two in
%our system, so we assume all examples are positive.
%
%Chen~\cite{chen1995bayesian} uses Bayesian inference to converge
%on a suitable grammar, given examples.
%
%There has been recent interest in approximate type inference.
%
%Pluquet et. al~\cite{marot2009fast} investigate heuristics
%to quickly infer types in dynamic programs.
%So does Milojkovi{\'c}
%\cite{milojkovic2016exploring}.
%Spasojevi{\'c} et. al~\cite{spasojevic2014mining}
%compare types across a cross section of projects to improve
%inference.
%
%Adamsen et. al~\cite{adamsen2016analyzing} verify test suite completeness using a hybrid approach of lightweight dependency analysis, static type checking and dynamic instrumentation.
%
%% Inference and Evolution of TypeScript Declaration Files
%% - they submit pull requests from their tool's output
%% https://cs.au.dk/~amoeller/papers/tstools/paper.pdf
%
%% Automatic TS annotations from JSON (including recursive types)
%% https://github.com/shakyShane/json-ts

\paragraph{Automatic annotations}
There are two common implementation strategies for automatic annotation tools. The first
strategy, ``ruling-out'' (for invariant detection), assumes all invariants are true 
and then use runtime analysis results to rule out
impossible invariants. The second ``building-up'' strategy (for dynamic type inference)
assumes nothing and uses runtime analysis results to build up invariant/type knowledge.

Examples of invariant detection tools include Daikon~\infercitep{ernst2001dynamically},
DIDUCE~\infercitep{hangal2002tracking}, and Carrot~\infercitep{pytlik2003automated}, and
typically enhance statically typed languages with more expressive types or contracts.
Examples of dynamic type inference include our tool, Rubydust \infercitep{An10dynamicinference},
JSTrace~\infercitep{saftoiu2010jstrace}, and TypeDevil~\infercitep{pradel2015typedevil},
and typically target untyped languages.

Both strategies have different space behavior with respect to representing
the set of known invariants.
The ruling-out strategy typically uses a lot of memory at the beginning,
but then can free memory as it rules out invariants. For example, if
\texttt{odd(x)} and \texttt{even(x)} are assumed, observing \texttt{x = 1}
means we can delete and free the memory recording \texttt{even(x)}.
Alternatively, the building-up strategy uses the least memory storing
known invariants/types at the beginning, but increases memory usage
as more the more samples are collected. For example, if we know
\texttt{x : Bottom}, and we observe \texttt{x = "a"} and \texttt{x = 1}
at different points in the program, we must use more memory to
store the union \texttt{x : String $\cup$ Integer} in our set of known invariants.

\paragraph{Daikon}
Daikon can reason about very expressive relationships between variables
using properties like ordering ($x < y$), linear relationships ($y = ax + b$),
and containment ($x \in y$). It also supports reasoning with ``derived variables''
like fields ($x.f$), and array accesses ($a[i]$).
%
Typed Clojure's dynamic inference can record heterogeneous data structures
like vectors and hash-maps, but otherwise cannot express relationships
between variables.

There are several reasons for this. The most prominent is that Daikon
primarily targets Java-like languages, so inferring simple type information
would be redundant with the explicit typing disciplines of these languages.
On the other hand, the process of moving from Clojure to Typed Clojure
mostly involves writing simple type signatures without dependencies
between variables. Typed Clojure recovers relevant dependent information
via occurrence typing~\infercitep{TF10}, and gives the option to manually annotate necessary
dependencies in function signatures when needed.

\paragraph{Reverse Engineering Programs with Static Analysis}
Rigi~\infercitep{muller1992reverse} analyzes
the structure of large software systems,
combining static analysis 
with a user-facing graphical environment to allow users to view and manipulate
the in-progress reverse engineering results.
We instead use a static type system as a feedback mechanism,
which forces more aggressive compacting of generated annotations.

Lackwit~\infercitep{o1997lackwit} uses static analysis to identify abstract 
data types in C programs. Like our work, they share representations between
values, except they use type inference with representations encoded as types.
Recursive representations are inferred via Felice and Coppos's
work on type inference with recursive types~\infercitep{cardone1991type},
where we rely on our imprecise ``squashing'' algorithms over incomplete runtime samples.

Soft Typing~\infercitep{CF91} uses static analysis to insert runtime checks into untyped
programs for invariants that cannot be proved statically. Our approach is instead to let
the user check the generated annotations with a static type system, with static type errors
guiding the user to manually add casts when needed.

\paragraph{Schema Inference}
\infercitet{baazizi2017schema}
infer structural properties of JSON data using a custom JSON schema format.
Their schema inference algorithm proceeds in two stages:
schema inference and schema fusion.
This resembles our collection and naive type environment construction phases.
There are slight differences between schema fusion and our approach.
Schema fusion upcasts heterogeneous array types to be homogeneous, where
we maintain heterogeneous vector types until a differently-sized
vector type is found in the same position.
We also support function types, which JSON lacks.
While they support nested data, they do not attempt to factor out common types as names
or create recursive types like our squashing algorithms.

\infercitet{discala2016automatic}
present a machine learning algorithm to translate denormalized
and nested data that is commonly found in NoSQL databases to traditional
relational formats used by standard RDBMS.
A key component is a schema generation algorithm which arranges related
data into tables via a matching algorithm which discovers related attributes.
Phases 1 and 2 of their algorithm are similar to our local and global
squashing algorithms, respectively, in that first locally accessible information
is combined, and then global information.
%TODO do they infer recursive schemas?
They identify groups of attributes that have (possibly cyclic) relationships.
%They choose a loose method of relating entities (soft functional dependencies) to compensate
%for data inconsistencies and to help users learn about their data via higher-level abstractions.
Where our squashing algorithms for map types are based on (sets of) keysets---on the 
assumption that related entities use similar keysets---they also join attributes
based on their similar values.
This enables more effective entity matching via equivalent attributes
with different names (e.g., ``Email'' vs ``UserEmail'').
Our approach instead assumes programs are somewhat internally consistent, and instead
optimizes to handle missing samples from incomplete dynamic analysis.

%TODO Inferring XML Schema Definitions from XML Data - Bex, Neven, Vansummeren

% Inference and Evolution of TypeScript Declaration Files
% - they submit pull requests from their tool's output
% https://cs.au.dk/~amoeller/papers/tstools/paper.pdf
\paragraph{Other Annotation Tools}
Static analyzers
for JavaScript
(TSInfer~\infercitep{kristensen2017inference}) and for Python (Typpete~\infercitep{typette18}
and PyType~\infercitep{pytype})
automatically annotate code with types.
PyType and Typpete inferred \texttt{nodes}
as \texttt{(? -> int)}
and \texttt{Dict[(Sequence, object)] -> int}, respectively---our tool 
infers it as \clj{[Op -> Int]} by also generating a compact recursive
type.
Similarly, a class-based translation of
inferred both \texttt{left} and \texttt{right}
fields
as \texttt{Any} by PyType, and as \texttt{Leaf} by Typpete---our tool
uses \clj{Op},
a compact recursive type containing \emph{both} \clj{Leaf} and \clj{Node}.
This is similar to our experience with TypeWiz in \Dchapref{infer:chapter:intro}.
(We were unable to install TSInfer.)

NoRegrets~\infercitep{noregrets2018} uses dynamic analysis to learn how a program
is used, and automatically runs the tests of downstream projects to
improve test coverage.
Their \emph{dynamic access paths} represented as
a series of \emph{actions} are analogous to our paths of path elements.

% distinguishes public/private API

% Python
% - MaxSMT-Based Type Inference for Python 3
%  - cites other python based projects
%  - https://link.springer.com/content/pdf/10.1007%2F978-3-319-96142-2_2.pdf
% - pytype
%  - static analysis to generate python annotations
%  - https://github.com/google/pytype
% - pyannotate
%   - dynamic analysis
%   - https://github.com/dropbox/pyannotate

% A Survey of Dynamic Analysis and Test Generation for JavaScript
%  - http://mp.binaervarianz.de/js_survey_2017.pdf
