%% For double-blind review submission
%\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
\documentclass[acmlarge]{acmart}%\settopmatter{printfolios=true}

%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format should change 'acmlarge' to
%% 'sigplan,10pt'.


%% Some recommended packages.
%\usepackage{booktabs}   %% For formal tables:
%                        %% http://ctan.org/pkg/booktabs
%\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%                        %% http://ctan.org/pkg/subcaption
\usepackage{listings}
\lstdefinelanguage{clojure}%
{morekeywords={*,*1,*2,*3,*agent*,*allow-unresolved-vars*,*assert*,*clojure-version*,*command-line-args*,%
*compile-files*,*compile-path*,*e,*err*,*file*,*flush-on-newline*,*in*,*macro-meta*,%
*math-context*,*ns*,*out*,*print-dup*,*print-length*,*print-level*,*print-meta*,*print-readably*,%
*read-eval*,*source-path*,*use-context-classloader*,*warn-on-reflection*,+,-,->,->>,..,/,:else,%
<,<=,=,==,>,>=,@,accessor,aclone,add-classpath,add-watch,agent,agent-errors,aget,alength,alias,%
all-ns,alter,alter-meta!,alter-var-root,amap,ancestors,and,apply,areduce,array-map,aset,%
aset-boolean,aset-byte,aset-char,aset-double,aset-float,aset-int,aset-long,aset-short,assert,%
assoc,assoc!,assoc-in,associative?,atom,await,await-for,await1,bases,bean,bigdec,bigint,binding,%
bit-and,bit-and-not,bit-clear,bit-flip,bit-not,bit-or,bit-set,bit-shift-left,bit-shift-right,%
bit-test,bit-xor,boolean,boolean-array,booleans,bound-fn,bound-fn*,butlast,byte,byte-array,%
bytes,cast,char,char-array,char-escape-string,char-name-string,char?,chars,chunk,chunk-append,%
chunk-buffer,chunk-cons,chunk-first,chunk-next,chunk-rest,chunked-seq?,class,class?,%
clear-agent-errors,clojure-version,coll?,comment,commute,comp,comparator,compare,compare-and-set!,%
compile,complement,concat,cond,condp,conj,conj!,cons,constantly,construct-proxy,contains?,count,%
counted?,create-ns,create-struct,cycle,dec,decimal?,declare,def,definline,defmacro,defmethod,%
defmulti,defn,defn-,defonce,defprotocol,defstruct,deftype,delay,delay?,deliver,deref,derive,%
descendants,destructure,disj,disj!,dissoc,dissoc!,distinct,distinct?,do,do-template,doall,doc,%
dorun,doseq,dosync,dotimes,doto,double,double-array,doubles,drop,drop-last,drop-while,empty,empty?,%
ensure,enumeration-seq,eval,even?,every?,false,false?,ffirst,file-seq,filter,finally,find,find-doc,%
find-ns,find-var,first,float,float-array,float?,floats,flush,fn,fn?,fnext,for,force,format,future,%
future-call,future-cancel,future-cancelled?,future-done?,future?,gen-class,gen-interface,gensym,%
get,get-in,get-method,get-proxy-class,get-thread-bindings,get-validator,hash,hash-map,hash-set,%
identical?,identity,if,if-let,if-not,ifn?,import,in-ns,inc,init-proxy,instance?,int,int-array,%
integer?,interleave,intern,interpose,into,into-array,ints,io!,isa?,iterate,iterator-seq,juxt,%
key,keys,keyword,keyword?,last,lazy-cat,lazy-seq,let,letfn,line-seq,list,list*,list?,load,load-file,%
load-reader,load-string,loaded-libs,locking,long,long-array,longs,loop,macroexpand,macroexpand-1,%
make-array,make-hierarchy,map,map?,mapcat,max,max-key,memfn,memoize,merge,merge-with,meta,%
method-sig,methods,min,min-key,mod,monitor-enter,monitor-exit,name,namespace,neg?,new,newline,%
next,nfirst,nil,nil?,nnext,not,not-any?,not-empty,not-every?,not=,ns,ns-aliases,ns-imports,%
ns-interns,ns-map,ns-name,ns-publics,ns-refers,ns-resolve,ns-unalias,ns-unmap,nth,nthnext,num,%
number?,odd?,or,parents,partial,partition,pcalls,peek,persistent!,pmap,pop,pop!,pop-thread-bindings,%
pos?,pr,pr-str,prefer-method,prefers,primitives-classnames,print,print-ctor,print-doc,print-dup,%
print-method,print-namespace-doc,print-simple,print-special-doc,print-str,printf,println,println-str,%
prn,prn-str,promise,proxy,proxy-call-with-super,proxy-mappings,proxy-name,proxy-super,%
push-thread-bindings,pvalues,quot,rand,rand-int,range,ratio?,rational?,rationalize,re-find,%
re-groups,re-matcher,re-matches,re-pattern,re-seq,read,read-line,read-string,recur,reduce,ref,%
ref-history-count,ref-max-history,ref-min-history,ref-set,refer,refer-clojure,reify,%
release-pending-sends,rem,remove,remove-method,remove-ns,remove-watch,repeat,repeatedly,%
replace,replicate,require,reset!,reset-meta!,resolve,rest,resultset-seq,reverse,reversible?,%
rseq,rsubseq,second,select-keys,send,send-off,seq,seq?,seque,sequence,sequential?,set,set!,%
set-validator!,set?,short,short-array,shorts,shutdown-agents,slurp,some,sort,sort-by,sorted-map,%
sorted-map-by,sorted-set,sorted-set-by,sorted?,special-form-anchor,special-symbol?,split-at,%
split-with,str,stream?,string?,struct,struct-map,subs,subseq,subvec,supers,swap!,symbol,symbol?,%
sync,syntax-symbol-anchor,take,take-last,take-nth,take-while,test,the-ns,throw,time,to-array,%
to-array-2d,trampoline,transient,tree-seq,true,true?,try,type,unchecked-add,unchecked-dec,%
unchecked-divide,unchecked-inc,unchecked-multiply,unchecked-negate,unchecked-remainder,%
unchecked-subtract,underive,unquote,unquote-splicing,update-in,update-proxy,use,val,vals,%
var,var-get,var-set,var?,vary-meta,vec,vector,vector?,when,when-first,when-let,when-not,%
while,with-bindings,with-bindings*,with-in-str,with-loading-context,with-local-vars,%
with-meta,with-open,with-out-str,with-precision,xml-seq,zero?,zipmap
},%
   sensitive,% ???
   alsodigit=-,%
   morecomment=[l];,%
   morestring=[b]"%
  }[keywords,comments,strings]%

\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{stringgray}{rgb}{0.4,0.4,0.4}
\definecolor{mymauve}{rgb}{0.58,0,0.82}
\definecolor{types}{rgb}{0.16, 0.32, 0.75}
%\definecolor{types}{rgb}{0.0, 0.75, 1.0}
\definecolor{turnstile}{rgb}{0.0, 0.5, 1.0}
%\definecolor{interop}{rgb}{0,0.8,0}
%\definecolor{interop}{rgb}{0.89, 0.44, 0.48}
\definecolor{interop}{rgb}{0.01, 0.75, 0.24}
\definecolor{invoke}{rgb}{0,0,0}

\lstset{ %
  upquote=true,
  language=clojure,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\ttfamily,       % the size of the fonts that are used for the code
  %numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  %frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  %captionpos=t,           % sets the caption-position to bottom
  %breaklines=true,        % sets automatic line breaking
  %breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
  literate= 
            %{\#\{*set\ }{{\textcolor{red}{\#\{}}}1
            %{\ set*\}} {{\textcolor{red}{\}}}}1
            {\{}{{\textcolor{mygray}{\{}}}1
            {\}} {{\textcolor{mygray}{\}}}}1
            {\{*map\ }{{\textcolor{red}{\{}}}1
            {\ map*\}} {{\textcolor{red}{\}}}}1
            {[}  {{\textcolor{mygray}{[}}}1
            {]}  {{\textcolor{mygray}{]}}}1     
            {(*anon\ }  {{\textcolor{mygray}{\#(}}}1
            {(}  {{\textcolor{mygray}{(}}}1
            {)}  {{\textcolor{mygray}{)}}}1     
            {:-}  {{\textcolor{types}{:-}}}1     
            {->}  {{\textcolor{mygray}{->}}}1     
            {[*vec\ }  {{\textcolor{red}{[}}}1
            {\ vec*]}  {{\textcolor{red}{]}}}1     
            {(*list\ }  {{\textcolor{red}{(}}}1
            {\ list*)}  {{\textcolor{red}{)}}}1     
            {(*interop\ }  {{\textcolor{interop}{(}}}1     
            {\ interop*)}  {{\textcolor{interop}{)}}}1     
            {(*typed\ }  {{\textcolor{types}{(}}}1     
            {\ typed*)}  {{\textcolor{types}{)}}}1     
            {(*invoke\ }  {{\textcolor{invoke}{(}}}1     
            {\ invoke*)}  {{\textcolor{invoke}{)}}}1     
}

\lstset{ %
  %backgroundcolor=\color{white},   % choose the background color
  %basicstyle=\footnotesize,        % size of fonts used for the code
  commentstyle=\color{mygray},    % comment style
  escapeinside={(*@}{@*)},
  %escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
  keywordstyle=\textbf,       % keyword style
  stringstyle=\color{stringgray}\textit,     % string literal style
}

\newcommand{\clj}[1]{\lstinline{#1}}
\newcommand{\java}[1]{\lstinline{#1}}
\newcommand{\rkt}[1]{\lstinline{#1}}
\newcommand\secref[1]{Section~\ref{#1}}
\newcommand\figref[1]{Figure~\ref{#1}}
\newcommand\egref[1]{Example~\ref{#1}}
\newcommand\lemref[1]{Lemma~\ref{#1}}


\usepackage{fancyvrb}
\usepackage{amsmath}
\usepackage{mathpartir}
\usepackage{alltt,fancyvrb}
\usepackage{mmm}
\usepackage{style}
\usepackage{tikz}
\usepackage{tikz-qtree}
\bibliographystyle{abbrvnat}
\include{bibliography.bib}

\newcommand{\cL}{{\cal L}}

\begin{document}

%\title{Development and study of Verification Tools for Clojure}
%\title{Typed Clojure in Practice}
%\title{Theory and Practice of Verification Tools for Clojure}
%\title{Theory, Practice, and Study of }
%\title{Thesis Proposal: Theory, Practice, and Study of Clojure Verification Techniques for Clojure}
\title{Thesis Proposal: Typed Clojure in Theory and Practice}

%% Author with single affiliation.
\author{Ambrose Bonnaire-Sergeant}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  %\position{Position1}
  \department{Computer Science}              %% \department is recommended
  \institution{Indiana University}            %% \institution is required
  %\streetaddress{Street1 Address1}
  %\city{City1}
  %\state{State1}
  %\postcode{Post-Code1}
  \country{USA}
}
\email{abonnair@indiana.edu}          %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
The Clojure programming language has existing tooling runtime and compile-time
verification.

This thesis argues Typed Clojure is sound and practical,
that we can improve usability with automatic annotations based on runtime observations,
and by repurposing this annotation technology to generate clojure.spec,
we can answer broad questions about how Clojure is used.

I will first present Typed Clojure, an optional type system for Clojure.
I will show a formal model of Typed Clojure that includes
key features like hash-maps, multimethods, Java interoperability, and occurrence typing,
and prove the model type sound.
Then, I will demonstrate that Typed Clojure's design is useful and corresponds to actual usage patterns
with an emperical study of real-world Typed Clojure usage in over 19,000 lines of code.

Next, I will present a tool that automatically generates Typed Clojure annotations based on observed
program behavior.
First, I will show a formal model of the tool, including its runtime instrumentation phase that
collects samples from a running program, and type reconstruction phase
that creates useful annotations from these samples.
Then, I will synthesize these ideas into a practical implementation that

\end{abstract}

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

The popularity of dynamically-typed languages in software
development, combined with a recognition that types often improve
programmer productivity, software reliability, and performance, has
led to the recent development of a wide variety of optional and
gradual type systems aimed at checking existing programs written in
existing languages.  These include  TypeScript~\cite{typescript} and Flow~\cite{flow} for
JavaScript, Hack~\cite{hack} for PHP, and mypy~\cite{mypy}
for Python among the optional systems, and Typed Racket~\cite{TF08}, Reticulated
Python~\cite{Vitousek14}, and GradualTalk~\cite{gradualtalk} among gradually-typed systems.\footnote{We
  use ``gradual typing'' for systems like Typed Racket with sound
  interoperation between typed and untyped code; Typed Clojure or
 TypeScript which don't
  enforce type invariants we describe as ``optionally typed''.}

One key lesson of these systems, indeed a lesson known to early
developers of optional type systems such as StrongTalk, is that type
systems for existing languages must be designed to work with the
features and idioms of the target language. Often this takes the form
of a core language, be it of functions or classes and objects,
together with extensions to handle distinctive language features.

\subsection{Typed Clojure}

We synthesize these lessons to present \emph{Typed Clojure}, an
optional type system for Clojure. 
%
Clojure is a dynamically
typed language in the Lisp family---built on the Java Virtual
Machine (JVM)---which has recently gained popularity as an alternative
JVM language.  It offers the flexibility of a Lisp dialect, including
macros, emphasizes a functional style via
immutable data structures, and provides
interoperability with existing Java code, allowing programmers to use
existing Java libraries without leaving Clojure.
%
Since its initial release in 2007, Clojure has been widely adopted for
``backend'' development in places where its support for parallelism,
functional programming, and Lisp-influenced abstraction is desired on
the JVM. As a result, there is an extensive base of existing untyped
programs whose developers can benefit from Typed Clojure,
an experience we discuss in this paper.

Since Clojure is a language in the
Lisp family, we apply the lessons of Typed Racket, an existing gradual type
system for Racket, to the core of Typed Clojure, consisting of an extended
$\lambda$-calculus over a variety of base types shared between all Lisp systems.
%
Furthermore, Typed Racket's \emph{occurrence typing} has proved
necessary for type checking realistic Clojure programs.

\begin{figure*}[t!]
  \normalsize
\begin{lstlisting}
(*typed ann pname [(U File String) -> (U nil String)] typed*)
(defmulti pname class)  ; multimethod dispatching on class of argument
(defmethod pname String [s] (*invoke pname (*interop new File s interop*) invoke*)) ; String case 
(defmethod pname File [f] (*interop .getName f interop*)) ; File case, static null check
(*invoke pname "STAINS/JELLY" invoke*) ;=> "JELLY" :- (U nil Str)
\end{lstlisting}
\caption{A simple Typed Clojure program (delimiters: {\color{interop}Java interoperation (green)}, 
  {\color{types}type annotation (blue)},
  {\color{invoke}function invocation (black)}, {\color{red}collection literal (red)}, {\color{mygray}other (gray)})}
\label{fig:ex1}
\end{figure*}


However, Clojure goes beyond Racket in many ways, requiring several
new type system features which we detail in this paper.
%
Most significantly, Clojure supports, and Clojure developers use,
\textbf{multimethods} to structure their code in extensible
fashion. Furthermore, since Clojure is an untyped language, dispatch
within multimethods is determined by application of dynamic predicates
to argument values. 
%
Fortunately, the dynamic dispatch used by multimethods has surprising
symmetry with the conditional dispatch handled by occurrence
typing. Typed Clojure is therefore able to effectively handle complex
and highly dynamic dispatch as present in existing Clojure programs. 

But multimethods are not the only Clojure feature crucial to type
checking existing programs. As a language built on the Java Virtual
Machine, Clojure provides flexible and transparent access to existing
Java libraries, and \textbf{Clojure/Java interoperation} is found in almost
every significant Clojure code base. Typed Clojure therefore builds in
an understanding of the Java type system and handles interoperation
appropriately. Notably, \texttt{null} is a distinct type in Typed Clojure,
designed to automatically rule out null-pointer exceptions.

An example of these features is given in
\figref{fig:ex1}. Here, the \clj{pname} multimethod dispatches
on the \clj{class} of the argument---for \clj{String}s,
the first method implementation is called, for \clj{File}s, the
second. The \clj{String} method calls
a \clj{File} constructor, returning a non-nil \clj{File} instance---the 
\clj{getName} method 
on \clj{File} requires a non-nil target, returning a nilable
type.  
%Typed Clojure offers an opt-in mode that
%resolves JVM overloading, avoiding expensive runtime reflective calls.

Finally, flexible, high-performance immutable dictionaries
are the most common Clojure data structure.
Simply treating them as uniformly-typed
key-value mappings would be insufficient for existing
programs and programming styles. Instead, Typed Clojure provides a
flexible \textbf{heterogenous map} type, in which specific entries can be specified. 

While these features may seem disparate, they are unified in important
ways. First, they leverage the type system mechanisms
inherited from Typed Racket---multimethods when using 
dispatch via predicates, Java interoperation for handling
\texttt{null} tests, and heterogenous maps using union types and
reasoning about subcomponents of data. Second,
they are crucial features for handling Clojure code in
practice. Typed Clojure's use in real Clojure deployments would not be
possible without effective handling of these three Clojure features. 

\subsection{Automatic Type Annotations}

% - set the scene for inferring types
%   - Typed Clojure
%   - optional/gradual typing requires annotations

We now address major usability flaw
for gradually and optionally typed languages:
writing type annotations is a manual process.

Take \texttt{vertices} (Figure~\ref{vertices}),
written in Clojure.
As is good style, it comes with a unit test.
Our goal is to \textit{generate} Typed Clojure~\cite{typed-clojure}
annotations
for this function, relieving most of the annotation
burden.

Our approach features several stages.
First, we \textit{instrument} top-level functions
%(Section \ref{instrument-TODO})
,
then run the unit tests and \textit{track}
how they are used at runtime
%(Section \ref{track-TODO})
.
At this point, we have a preliminary
annotation:
%
\begin{Verbatim}
(ann vertices ['{:op ':node, :left '{:op ':leaf, :val Int}, :right '{:op ':leaf, :val Int}} -> Int]})
\end{Verbatim}
%
This type is too specific---trees are recursively
defined---we \textit{squash} types to be
recursive from example unrollings%(Section \ref{recursive-TODO})
:

\begin{Verbatim}
(defalias NodeLeaf 
  (U '{:op ':node :left NodeLeaf :right NodeLeaf}
     '{:op ':leaf :val Int}))
(ann vertices [NodeLeaf -> Int])
\end{Verbatim}
%
%\begin{Verbatim}
%(declare Node Leaf)
%(defalias NodeLeaf (U Node Leaf))
%(defalias Node 
%  '{:op ':node :left NodeLeaf :right NodeLeaf})
%(defalias Leaf '{:op ':leaf :val int})
%(ann verbatim [NodeLeaf -> Int]})
%\end{Verbatim}
%
If \texttt{NodeLeaf} is used in multiple positions
in the program, local recursive types are redundant.
In this paper, we name and \textit{merge} recursive
types, reusing them in annotations.
%
\begin{Verbatim}
(ann vertices [NodeLeaf -> Int])
(ann sum-tree [NodeLeaf NodeLeaf -> NodeLeaf])
\end{Verbatim}
%
If minor variants of the recursive types occur
across a program,
we use \textit{optional} entries~\cite{typed-clojure}
to reduce redundancy.
%(Section \ref{optional-merge-TODO})
\begin{Verbatim}
(defalias NodeLeaf 
  (U '{:op ':node :left NodeLeaf :right NodeLeaf}
     (HMap :mandatory {:op ':leaf :val Int}
           :optional {:label Str})))
\end{Verbatim}

After inserting these annotations, we can run the
type checker over them to check their usefulness.
We found annotations to be readable and minimize
redundancy compared to hand-written annotations
%(Section \ref{experiment1})
.
Minimal changes were needed to successfully type check
functions with the generated annotations,
mostly consisting of local function and loop annotations,
and renaming of type aliases.
%(Section \ref{experiment2})
Generating and running \textit{tests} improved the quality
of type annotations by exercising more paths through the
program. %(Section \ref{experiment3})

\subsection{Investigation}

\section{Thesis Statement}

My thesis statement is:

\begin{quotation}
Typed Clojure is a sound and practical optional type system for Clojure whose annotation burden is partially-automatable, and this automation technology can be repurposed to further reveal how Clojure is used in real projects.
%Typed Clojure is sound, practical, and its annotation burden is partially-automatable,
%and we can repurpose this annotation technology to
%answer broad questions about how Clojure is used.
%Typed Clojure is a well-founded and practical optional type system for Clojure
%whose useability can be improved by
%developing a tool to automatically generate type annotations,
%and, by repurposing this tool, we can study general Clojure idioms and practices across
%hundreds of projects by generating, running, and exercising clojure.spec runtime specifications.
\end{quotation}

I will support this thesis statement with the following:

\begin{itemize}
  \item \emph{Typed Clojure is sound} We formalize Typed Clojure, including
    its characteristic features like hash-maps, multimethods, and Java interoperability,
    and prove the model type sound.
  \item \emph{Typed Clojure is practical} We present an emperical study of real-world Typed Clojure usage
    over 19,000 lines of code, showing its features correspond to actual usage patterns.
  \item \emph{Typed Clojure's annotation burden is partially-automatable} We formalize and implement a tool to generate
    type annotations, and emperically study the manual changes needed for the generated annotations
    to pass type checking.
  %\item \emph{clojure.spec model and Automatic Annotations} 
  %\item \emph{Repurpose automation technology}
  \item \emph{Repurpose automation technology}
    We describe how to automatically generate clojure.spec annotations (``specs'') for existing programs.
    We present a formal model of clojure.spec (an existing and popular runtime verification tool for Clojure)
    and implement the model in Redex.
  \item \emph{Further reveal how Clojure is used in real projects}
    We conduct a study of general Clojure idioms and practices by generating, enforcing, and exercising specs
    across hundreds of projects, as well as justifying design choices in Typed Clojure's type system,
    clojure.spec's features, and our automatic annotation tool.
\end{itemize}

\section{Technical Overview}

\input{track}

\section{Related Work}

% Cite a few of the early papers here.
%http://www.cs.washington.edu/research/projects/cecil/www/pubs/
\paragraph{Multimethods} 
Millstein and collaborators present a sequence of
systems~\cite{Chambers:1992:OMC,Chambers:1994:TMM,MS02} with statically-typed multimethods
and modular type checking.  In contrast to Typed Clojure, in these
system methods declare the types of arguments that they expect which
corresponds to exclusively using \clj{class} as the dispatch function
in Typed Clojure. However, Typed Clojure does not attempt to rule out
failed dispatches.

% one sentence
% TC based on TR, already covered

\paragraph{Occurrence Typing} 
Occurrence typing~\cite{TF08,TF10} extends the type 
system with a \emph{proposition environment} that represents 
the information on the types of bindings down conditional branches.
These propositions are then used to update the types associated
with bindings in the \emph{type environment} down branches
so binding occurrences are given different types 
depending on the branches they appear in, and the conditionals
that lead to that branch.

% What's diff about TC from the related work
% small summary for deisel....
% - diesel supports x
%- - calculus supports some subset of x
% we support y, which covers most of x but also foo

% eg. multiple dispatch
%     nominal vs structural

% eg. run abritrary metaprogramming over dispatch in CLOS
%  more expressive

% type systems for mm or rows
% rows vs HMap
% - no poly in HMap
% - based on subtyping
% - rows based on polymorphism

\paragraph{Record Types} Row polymorphism~\cite{Wand89typeinference,CM91,HP91}, used
in systems such as the OCaml object system, provides many of the
features of HMap types, but defined using universally-quantified row
variables. HMaps in Typed Clojure are instead designed to be used with
subtyping, but nonetheless provide similar expressiveness, including
the ability to require presence and absence of certain keys. 

Dependent JavaScript~\cite{Chugh:2012:DTJ} can track similar
invariants as HMaps with types for JS objects. They must deal with
mutable objects, they feature refinement types and strong updates to
the heap to track changes to objects.

TeJaS~\cite{TeJaS}, another type system for JavaScript,
also supports similar HMaps, with the ability to
record the presence and absence of entries, but lacks a compositional
flow-checking approach like occurrence typing.

Typed Lua~\cite{Maidl:2014:TLO} has \emph{table types} which track
entries in a mutable Lua table.  Typed Lua changes the dynamic
semantics of Lua to accommodate mutability: Typed Lua raises a runtime
error for lookups on missing keys---HMaps consider lookups on missing
keys normal.

\paragraph{Java Interoperability in Statically Typed Languages}
Scala~\cite{OCD+} has nullable references for compatibility with Java.
Programmers must manually check for
\java{null} as in Java to avoid null-pointer exceptions. 


\paragraph{Other optional and gradual type systems}
%In addition to Typed Racket, 
Several other gradual type
systems have been developed for existing
dynamically-typed languages.  Reticulated Python~\cite{Vitousek14} is
an experimental gradually typed system for Python, implemented as a
source-to-source translation that inserts dynamic checks at language
boundaries and supporting Python's first-class object system. 
Clojure's nominal classes avoids the need to support
first-class object system in Typed Clojure, however HMaps offer an alternative to
the structural objects offered by Reticulated. Similarly,
Gradualtalk~\cite{gradualtalk} offers gradual typing for Smalltalk,
with nominal classes.

Optional types
%, requiring less implementation effort and avoiding
%runtime cost, 
have been  adopted in industry, including Hack~\cite{hack}, and Flow~\cite{flow} and
TypeScript~\cite{typescript}, two extensions of JavaScript. These
systems  support  limited forms of occurrence typing,
and do not include the other features we
present.

%  \item GradualTalk
%  \item Flow
%\end{itemize}

\paragraph{Automatic annotations}
There are two common implementation strategies for such tools. The first
strategy, ``ruling-out'' (for invariant detection), assumes all invariants are true 
and then use runtime analysis results to rule out
impossible invariants. The second ``building-up'' strategy (for dynamic type inference)
assumes nothing and then uses runtime analysis results to build up invariant/type knowledge.

Examples of invariant detection tools include Daikon \cite{Ernst06thedaikon},
DIDUCE \cite{hangal2002tracking}, and Carrot \cite{pytlik2003automated}, and
typically enhance statically typed languages with more expressive types or contracts.
Examples of dynamic type inference include Rubydust \cite{An10dynamicinference},
JSTrace \cite{saftoiu2010jstrace}, and TypeDevil \cite{pradel2015typedevil},
and typically target untyped languages.

Both strategies have different space behavior with respect to representing
the set of known invariants.
The ruling-out strategy typically uses a lot of memory at the beginning,
but then can free memory as it rules out invariants. For example, if
\texttt{odd(x)} and \texttt{even(x)} are assumed, observing \texttt{x = 1}
means we can delete and free the memory recording \texttt{even(x)}.
Alternatively, the building-up strategy uses the least memory storing
known invariants/types at the beginning, but increases memory usage
as more the more samples are collected. For example, if we know
\texttt{x : Bottom}, and we observe \texttt{x = "a"} and \texttt{x = 1}
at different points in the program, we must use more memory to
store the union \texttt{x : String $\cup$ Integer} in our set of known invariants.

\paragraph{Daikon}
Daikon can reason about very expressive relationships between variables
using properties like ordering ($x < y$), linear relationships ($y = ax + b$),
and containment ($x \in y$). It also supports reasoning with ``derived variables''
like fields ($x.f$), and array accesses ($a[i]$).

Typed Clojure's dynamic inference can record heterogeneous data structures
like vectors and hash-maps, but otherwise cannot express relationships
between variables.

There are several reasons for this. The most prominent is that Daikon
primarily targets Java-like languages, so inferring simple type information
would be redundant with the explicit typing disciplines of these languages.
On the other hand, the process of moving from Clojure to Typed Clojure
mostly involves writing simple type signatures without dependencies
between variables. Typed Clojure recovers relevant dependent information
via occurrence typing, and gives the option to manually annotate necessary
dependencies in function signatures when needed.

% Inference and Evolution of TypeScript Declaration Files
% - they submit pull requests from their tool's output
% https://cs.au.dk/~amoeller/papers/tstools/paper.pdf

\section{Research Plan and Timeline}

I plan to publish the following papers:

\begin{itemize}
  \item \textbf{ESOP 2016 - Published} \emph{Practical Optional Types for Clojure}.
    A paper that provides a formal model of Typed Clojure and presents an emperical
    study of usage patterns.
  \item \textbf{Spring 2018} \emph{Squash the work: Automatic annotations for Typed Clojure}
		A paper that presents a tool capable of generating Typed Clojure and clojure.spec
		annotations based on runtime observations of the program.
		We present a formal model of the tool, as well as several manual experiments in how
		accurate our annotations are.
    We conduct a larger study by sourcing several hundred projects and automatically
    generating types and specs, then using these annotations to answer various questions
    about how clojure.spec and Clojure is used.
\end{itemize}

My timeline is:

\begin{itemize}
  \item \textbf{[June-August 2018]} Finish automatic annotation experiments, including
    manually measuring the difference between automatically generated annotations and
    fully type-checked annotations, and large scale generation of specs for
    hundreds of projects.
  \item \textbf{[September-November 2018]} Write paper (PLDI submission)
  \item \textbf{[November 2018-May 2019]} Write dissertation
  \item \textbf{[June 2019]} Defend
\end{itemize}

\bibliography{bibliography}


\end{document}
